Re-examining cross-cultural similarity judgments using lexical co-occurrence

### 2023-01-31 code/4_paper/manuscript/manuscript.tex
manuscript edits

By virtue of discrete words and grammatical features, language provides
a categorical partition of our continuous experiences. By measuring the
similarity between words (as commonly done with lexical co-occurrence
models), it may be possible to model the shape of similarity space in
order to capture cross-linguistic and cross-cultural variation and
provide a more general answer to the question of how language influences
our conception of the world.

Taxonomic and thematic similarity provide a convenient entry point to
broader debates about cross-linguistic and cross-cultural variation in
notions of similarity. Taxonomic categorization is based on the
similarity of attributes, for example, similar perceptual properties,
like shared color or shape, among objects. In contrast, thematic
categorization is based on causal, spatial, and temporal relationships
among objects (Markman \& Hutchinson, 1984).

@@

Preferences for taxonomic and thematic similarity vary across cultures.
Chiu (1972) found that Chinese children (9-10 years old) are more likely
to choose thematic matches in a picture triad task (shown, e.g., images
of an ant, a bee, and honey) than their American counterparts. These
cross-cultural differences are also observed in novel object
categorization, with Chinese participants preferring to group by family
resemblance across multiple features and Americans preferring a
single-feature rule (Norenzayan, Smith, Kim, \& Nisbett, 2002). The
authors link these differences to tendencies toward analytic processing
in Western cultures, which emphasizes objects and their properties, and
holistic processing in East Asian cultures, which emphasizes
relationships between objects and their context (see also Nisbett
(2003)). In related work, East Asian participants show a higher level of
sensitivity to context than their Western counterparts when reproducing
drawings from memory (Ji, Peng, \& Nisbett, 2000); visually exploring
naturalistic scenes (Chua, Boland, \& Nisbett, 2005); describing scenes
(Masuda \& Nisbett, 2001); and in explaining the causes of ambiguous
behaviors (Choi, Nisbett, \& Norenzayan, 1999).

Ji, Zhang, \& Nisbett (2004) asked whether differences in analytic and
holistic processing are driven by language, culture, or a combination of
these. They presented Chinese and European American adults with a word
triad task, and found a preference for thematic matching among Chinese
participants compared to European Americans. They found that test
language contributed to this difference, but did not fully explain it:
Chinese participants showed a stronger thematic preference when tested
in Mandarin, but still showed a thematic preference when tested in
English. Ji et al.~conclude that culture (independent of language) leads
to different styles of reasoning about similarity, whereas language
serves as a ``tuning'' mechanism operating within a culturally-specific
style, by activating representations corresponding to the language being
used.\footnote{But note that this work did not include a manipulation to
  prevent participants from engaging in e.g., covert naming in a
  language other than the test language.}

This view ascribes cross-cultural differences in similarity judgment to
variation in the conceptualization of similarity itself. Alternatively,
these judgments could be shaped by cross-cultural differences in the
statistics of the environment, and accordingly the content of everyday
experiences. Perhaps when confronted with the triad task, participants
from all cultures follow the same process for reasoning about
similarity, but rely on language or culture-specific inputs to this
process. If we observe a difference in categorization between East Asian
and Western participants, it could be that members of both groups use
the same procedure (considering similarity that is influenced by both
taxonomic and thematic relations), but the inputs to this procedure
differ between cultures, with East Asian participants exposed to more
instances of thematic similarity than their Western counterparts.

@@

While co-occurrences in experience are difficult to measure,
co-occurrence in language can provide a rough proxy--and indeed,
language may afford many of the ``experiences'' that people have with
infrequently encountered items, like cows or helicopters. In this study,
we use co-occurrence in language as a proxy of cultural experience. This
is a generalizing assumption, and while no linguistic corpus can be
expected to fully capture a culture, using this proxy as a model of
culture provides a conservative test of the hypothesis that differences
in the input, rather than the process, of reasoning produces cultural
variation in this task. This means that if such a model does work -- if
linguistic co-occurrence can predict cultural variation in this
similarity task without building differences in the decision process
into the model -- it is relatively strong evidence that language alone
(perhaps as a more easily observed source of information on broader
cultural and ecological variation) may account for these cross-cultural
differences in reasoning.

Indeed, previous work suggests that using lexical co-occurrence as a
proxy can be useful in thinking about similarity reasoning. Natural
language processing tasks have found that lexical co-occurrence is a
good predictor for performance in similarity judgment. Griffiths,
Steyvers, \& Tenenbaum (2007) showed that a model that takes
word-document co-occurrence as input can be used to predict word
association and the effects of semantic association on a variety of
linguistic tasks. Additionally, lexical semantic models that are
constructed using lexical co-occurrence (as opposed to annotated
relations) have been shown to perform well on predicting human judgments
about similarity between word pairs that are thematically or
taxonomically related (Rohde, Gonnerman, \& Plaut, 2006).

Our study uses cosine distances of fastText word vectors as a measure of
lexical co-occurrence\footnote{We also carried out our analysis using
  raw lexical co-occurrences and obtained similar results.}. fastText is
a system that uses lexical co-occurrence information to generate a
vector representing each word in its lexicon (Mikolov, Grave,
Bojanowski, Puhrsch, \& Joulin, 2018).

Word embeddings like fastText have been demonstrated to be good
predictors for similarity judgments (Liu, Feng, Wu, Chan, \& Fulton
(2019), Jatnika, Bijaksana, \& Suryani (2019)). fastText has also been
shown to be sensitive towards cultural effects on word meanings:
Thompson \& Lupyan (2020) showed that the distribution of semantic
meaning clusters generated by fastText trained on language-specific
corpora correlates to the cultural, historical and geographical
distances of such languages languages. This is a proof of concept for
our use of fastText in different languages with varying levels of
relatedness.

@@

Ji et al. (2004) suggest that the notion of similarity is what varies
across cultural groups, but it is possible that similarity is regarded
in the same way and it is actually the input to this similarity judgment
that varies across cultural contexts. It may be possible to gain
traction on potential mechanisms by examining whether variation in
similarity judgments co-vary with environmental statistics that differ
across cultural and linguistic contexts.

In this study, we ask whether variation in lexical co-occurrence can
account for the cross-context (culture and language) differences
observed in how people evaluate similarity. First, we attempt to
replicate Ji et al. (2004) by running a study comparing English speakers
in the US and Mandarin speakers in mainland China. Second, we evaluate
whether the differences observed between English and Mandarin speakers'
similarity judgments also extend to comparisons between English and
Vietnamese, by running a similar study between English speakers in the
US and Vietnamese speakers in Vietnam. Vietnam is a Southeast Asian
country that borders China and is historically greatly influenced by
Chinese culture (Hui, 2002). Therefore, it serves as a suitable cultural
context to investigate whether the claim made by Ji et al. (2004) and
previous studies, that Eastern and Western cultures have different
notions of similarity, extends beyond mainland China.

We then ask to what extent similarity judgments show cross-context
differences that align with co-occurrence patterns in participants'
languages. We would only see a main effect of cultural context if
differing notions of similarity drive these cross-cultural differences.
Meanwhile, if we do see variation in responding trial-to-trial that
tracks with lexical co-occurrence statistics (as a rough proxy combining
both linguistic and cultural context), this suggests that cross-cultural
notions of similarity might be similar, and it is environmental
statistics that guide the differences in similarity judgments.

To preview our results, we find that, while we do not find the predicted
Eastern/Western contrast in notions of similarity in our Vietnam
extension, we do find language-specific co-occurrence can indeed explain
cross-context differences in similarity judgments. In contrast to
previous accounts, by which culture induces differing conceptions of
similarity (varying between taxonomic and thematic), these findings
provide an alternative, and more specific, account by which language may
explain these cross-cultural differences without invoking variation in
notions of similarity.

There are several important limitations of our approach. While we
discuss cross-cultural variability at the level of countries or larger
world areas, these are not cultural monoliths. For convenience, we
operationalize culture at the level of country, based on where
participants were raised. It is an open question whether performance in
our participant populations (of relatively young and well-educated
adults) is representative of the broader country. This is especially
true for societies with substantial ethnic and cultural variation such
as the US. We expect that our data is likely to underestimate variation
both within and between the countries we sample from.

Language, culture, cognition, and individual experiences are intertwined
in complex causal relationships: cultural features (farming) can lead to
differences in individuals' experiences (seeing cows and grass) and in
turn these can influence language (talking more about cows and grass).
But cultural features (a thematic orientation) could also more directly
cause individuals to talk differently about the same experiences
(mentioning what cows eat rather than what other animals cows are like).
In this study, we measure language and its relation to cross-cultural
differences in categorization, but these relations test only the
plausibility of a language-based account; they cannot establish the
direction of causality.

@@

We excluded 8 US participants, 62 Vietnam participants, and 16 China
participants who did not answer 2 or more attention checks correctly. We
followed 4 exclusion criteria that aim to retain only participants who
are influenced by one culture: (1) non-native speakers of English and
Vietnamese, respectively, (2) fluent in at least one of the other two
study languages (Vietnamese for US participants, English for Vietnamese
participants and Chinese participants), (3) have lived outside of the
test country (US, Vietnam, or China) for more than two years, and (4)
have significant international experience (more than 6 international
experiences of 2 days or longer.) We did not use a particular criterion
for a language if it would exclude 25\% or more of any one sample. In
this round of exclusion, we excluded 73 US, 27 Vietnam participants, and
35 China participants. After these exclusions, the US sample included
119 participants (30M, 84F, 3 non-binary, 2 other), with mean age = 22.2
(SD = 8.15) and median age = 20. The Vietnam sample included 110
participants (34M, 71F, 5 other), with mean age = 22.21 (SD = 5.81) and
median age = 21. The China sample included 149 participants (61M, 87F, 1
other), with mean age = 23.1 (SD = 3.65) and median age = 23.\footnote{A
  table summarizing number of participants lost at each round of
  exclusions is included in the Appendix.}

Notably, we lost a majority of our Vietnam participants (more than 60\%)
in the attention check exclusion. One reason we suspect why this might
have happened is because our Vietnam participants are less familiar with
research surveys and attention check questions, and thus might have
thought too much about the attention check questions.\footnote{We
  carried out an exploratory analysis where we used a less stringent
  attention check exclusion, and did not find any difference in
  significant results when compared with the main analysis (see
  Appendix).}

\hypertarget{stimuli}{%
\subsection{Stimuli}\label{stimuli}}

We adapted stimuli from previous studies to create a set of test triads
consisting of a cue, with one thematic and one taxonomic match option.
For example, ``cow,'' ``grass,'' and ``chicken,'' where ``cow'' is the
cue, ``grass'' is the thematic match, and ``chicken'' the taxonomic
match. We included 105 such triads, a superset including triads pulled
from supplemental information and in-text examples across the
literature, and others that we adapted or created . We selected triads
on the basis of cultural familiarity in the US, Vietnam, and China. The
triads were originally in English; they were translated to Vietnamese
and Mandarin by a fluent bilingual speaker in each language. The
translations were then checked for accuracy after backtranslation to
English by another fluent bilingual in each language who was naive to
the original English versions. --\textgreater{}

Each participant completed all 105 triads in sets of 21 trials at a time
(10 test triads, 10 filler triads, and 1 attention check per page), by
selecting the match most related to the cue (``Which thing is most
closely related to the bolded item?'' or ``Thứ nào liên quan nhất với
thứ được in đậm?''). The test triads were presented with 105 filler
triads mixed in, to obscure the taxonomic-thematic two-answer forced
choice structure of the test stimuli and reduce the likelihood that
participants would become aware of the design. The filler triads were
groups of three semantically related words, but where the match options
were not distinguished by thematic vs.~taxonomic similarity, e.g., cue
``bird'' with match options ``lizard'' and ``toad.'' Additionally, we
included 10 attention check trials, which were formatted like the test
and filler triads but included an instruction instead of a cue item,
e.g., ``Choose wife'' with match options ``wife'' and ``husband.'' In
total, each participant completed 210 similarity judgments and 10
attention check questions, with triads presented in randomized orders
that varied between subjects.

@@

To give an intuition for our model, consider again the cow-grass-chicken
triad: we retrieved word vectors for ``cow'' and ``grass'', and
calculate the cosine distance between these vectors. Similarly, we
retrieved vectors for ``cow'' and ``chicken'' and calculate the cosine
distance between them. Our similarity prediction is then inversely
proportional to the cosine distance of these pairs. This is because a
larger cosine distance means the word vectors are further apart, and
thus the words are less similar. For example, if the cosine distance of
thematic cow-grass is 0.7 and the cosine distance of taxonomic
cow-chicken is 0.3, then our model predicts, correspondingly, that 30\%
of responses to the triad will be grass, and the other 70\% chicken. We
then use a mixed-effects regression to evaluate how well each corpus
model predicts participants' similarity judgments, across triads and
languages.

@@

We use the fastText pre-trained models of English, Mandarin, and
Vietnamese in Grave, Bojanowski, Gupta, Joulin, \& Mikolov (2018). These
models are trained on Common Crawl and Wikipedia using We distribute
pre-trained word vectors for 157 languages, trained on Common Crawl and
Wikipedia using fastText. These models were trained using a Continuous
Bag of Words (CBOW) with position-weights and a window of size 5. The
models use character n-grams of length 5 and 10 negative examples.From
the aforementioned models, we retrieve the word vectors (dimension 300)
for each word we are interested in.

@@

From the word vectors, we calculated the cosine distance between each
cue-thematic match (thematic cosine distance) and cue-taxonomic match
(taxonomic cosine distance), using the spatial.distance.cosine function
from the SciPy package (Virtanen et al., 2020). We then calculated the
thematic cosine distance proportion as thematic cosine distance over the
sum of taxonomic cosine distance and thematic cosine distance. We did
this for all three corpora. We were able to obtain predictions for all
triads in all languages.

@@

\hypertarget{do-we-extend-previous-work-reporting-a-preference-for-taxonomic-matching-in-the-us-and-thematic-matching-in-asia}{%
\subsection{1. Do we extend previous work reporting a preference for
taxonomic matching in the US and thematic matching in
Asia?}\label{do-we-extend-previous-work-reporting-a-preference-for-taxonomic-matching-in-the-us-and-thematic-matching-in-asia}}

@@

To test for cross-context differences in similarity judgments between
the countries, we ran a mixed-effects logistic regression predicting
triad responding (taxonomic or thematic) with country (US, China, or
Vietnam) as a fixed effect. As random effects, we included an intercept
per subject and one per triad, as well as by-triad random slopes for
country to account for variation in the country effect across triads. In
R syntax, the model is: response \textasciitilde{} country + (1
\textbar{} subject) + (country \textbar{} triad).

@@

\hypertarget{can-the-differences-in-similarity-judgments-between-english-mandarin-and-vietnamese-speakers-be-explained-by-variation-in-lexical-statistics}{%
\subsection{2. Can the differences in similarity judgments between
English, Mandarin and Vietnamese speakers be explained by variation in
lexical
statistics?}\label{can-the-differences-in-similarity-judgments-between-english-mandarin-and-vietnamese-speakers-be-explained-by-variation-in-lexical-statistics}}

\hypertarget{is-responding-in-each-cultural-context-predicted-by-lexical-statistics}{%
\subsubsection{Is responding in each cultural context predicted by
lexical
statistics?}\label{is-responding-in-each-cultural-context-predicted-by-lexical-statistics}}

To test whether variation in lexical statistics can explain differences
in similarity judgments between US and Vietnam participants, we compare
logistic mixed-effects regression models fit to the thematic responding
data from each country separately. We first ask how well each corpus
model (English, Vietnamese, or Mandarin) predicts similarity judgments
by speakers of the corresponding language (US, Vietnam, or China). To do
this, we use a mixed-effects logistic regression to predict triad
responses (0=taxonomic or 1=thematic) with corpus prediction (proportion
of cosine distance) as a fixed effect and participant and triad as
random effects.

We found that all corpora are significant predictors of all cultural
context responding, with p \textless{} 0.05 and \(\beta\) from -8.69 to
-2.4. (For a full report, see Appendix.)

\hypertarget{is-responding-in-each-cultural-context-best-predicted-by-the-corresponding-corpuss-lexical-statistics-as-opposed-to-the-other-two-corpora}{%
\subsubsection{Is responding in each cultural context best predicted by
the corresponding corpus's lexical statistics, as opposed to the other
two
corpora?}\label{is-responding-in-each-cultural-context-best-predicted-by-the-corresponding-corpuss-lexical-statistics-as-opposed-to-the-other-two-corpora}}

Next, we directly compare the corpus models by including both as fixed
effects in three mixed-effect regressions (predicting US, Vietnam and
China responding) with the same random effects as above. In R syntax,
the model is: response \textasciitilde{} corpus\_English +
corpus\_Vietnamese + corpus\_Mandarin + (1\textbar triad) +
(1\textbar subject).

For US responding: English (EN), Vietnamese (VI), and Mandarin (ZH)
corpus are all significant predictors. EN corpus: \(\beta\) = -6.96,
\(\chi^2\)(1) = 16.57, p \textless{} .001. VI corpus: \(\beta\) = -2.22,
\(\chi^2\)(1) = 3.73, p = 0.054. ZH corpus: \(\beta\) = -3.18,
\(\chi^2\)(1) = 3.37, p = 0.066.

For Vietnam responding: English (EN), Vietnamese (VI), and Mandarin (ZH)
corpus are all significant predictors. EN corpus: \(\beta\) = -2.98,
\(\chi^2\)(1) = 2.58, p = 0.108. VI corpus: \(\beta\) = -2.75,
\(\chi^2\)(1) = 4.84, p = 0.028. ZH corpus: \(\beta\) = -4.39,
\(\chi^2\)(1) = 5.49, p = 0.019.

For China responding: English (EN), Vietnamese (VI), and Mandarin (ZH)
corpus are all significant predictors. EN corpus: \(\beta\) = -3.32,
\(\chi^2\)(1) = 7.18, p = 0.007. VI corpus: \(\beta\) = -1.59,
\(\chi^2\)(1) = 3.63, p = 0.057. ZH corpus: \(\beta\) = -5.31,
\(\chi^2\)(1) = 17.69, p \textless{} .001.

We observed some level of language specificity from this analysis. The
English corpus is the best predictor for US responding, and the Mandarin
corpus is the best predictor for China response. While this is not the
case with the Vietnamese corpus and the Vietnam responding, the
Vietnamese corpus is still a significant predictor for the Vietnam
responding (Figure 2).

@@

However, we found that language specificity alone does not explain our
results. We used an ANOVA to compare the model with only the
corresponding corpus, and the model with all 3 corpora for each cultural
context. We found that in all three cases (US, China, Vietnam), adding
the other two corpora produces a significantly better fit than the
identical model without the additional corpora, and only the
corresponding corpus included as a predictor (US response: \(\chi^2\)(2)
= 7.93, p = 0.019; Vietnam response: \(\chi^2\)(2) = 13.72, p = 0.001;
China response: \(\chi^2\)(2) = 10.56, p = 0.005).

>>>>

@@

The way we partition our continuous experiences of the world into
categories, and which kind of similarity relationships we privilege to
organize such partitions, vary across cultures. One way this
cross-cultural difference has been observed is with a task comparing
preference for taxonomic and thematic similarity between East Asian and
Western participants.\footnote{Taxonomic categorization is based on the
  similarity of attributes, for example, similar perceptual properties,
  like shared color or shape, among objects.In contrast, thematic
  categorization is based on causal, spatial, and temporal relationships
  among objects (Markman \& Hutchinson, 1984).} Ji, Zhang, \& Nisbett
(2004) found that in a word triad task (choose two out of three words
that are more related to one another), Chinese participants preferred
thematic matching more compared to European Americans. In a pictorial
version of this task, Chiu (1972) found that Chinese children (9-10
years old) are also more likely to choose thematic matches compared to
their American counterparts. This cross-cultural difference is also
observed in novel object categorization, with Chinese participants
preferring to group by family resemblance across multiple features and
Americans preferring a single-feature rule (Norenzayan, Smith, Kim, \&
Nisbett, 2002).

Previous research has linked the observed difference in similarity
judgment to related work showing tendencies toward analytic processing
(emphasizing objects and their properties) in Western cultures and
holistic processing in East Asian cultures (emphasizing relationships
between objects and their context) (see also Nisbett (2003)).
Cross-cultural difference in similarity judgment is thus connected to
well-documented work showing East Asian participants exhibiting a higher
level of sensitivity to context than their Western counterparts when
reproducing drawings from memory (Ji, Peng, \& Nisbett, 2000); visually
exploring naturalistic scenes (Chua, Boland, \& Nisbett, 2005);
describing scenes (Masuda \& Nisbett, 2001); and in explaining the
causes of ambiguous behaviors (Choi, Nisbett, \& Norenzayan, 1999). This
view ascribes cross-cultural differences in similarity judgment to
variation in how each culture (or group of culture, i.e.~East Asian vs
Western) \textbf{conceptualize} similarity -- with East Asians
preferring thematic relations more than Westerners.

Alternatively, these judgments could be shaped by cross-cultural
differences in the \textbf{inputs} to similarity judgment, namely, the
statistics of the environment, and accordingly the content of everyday
experiences. Perhaps when faced with the triad task, participants from
all cultures follow the same process for conceptualizing similarity, but
rely on language or culture-specific inputs to this process. If we
observe a difference in categorization between East Asian and Western
participants, it could be that members of both groups use the same
procedure (considering similarity that is influenced by both taxonomic
and thematic relations), but the inputs to this procedure differ between
cultures, with East Asian participants exposed to more instances of
thematic similarity than their Western counterparts. We might also
expect that both conceptualization and inputs of similarity play a role
in driving cross-cultural differences -- East Asian participants may
both be exposed to more instances of thematic similarity, and prefer to
conceptualize similarity as thematic relations more so than Westerners.

@@

To test alternative hypotheses involving variation of inputs to
similarity judgment, we need a way to operationalize variation in
exposure to thematic vs taxonomic relations. While this metric is
difficult to directly measure, language statistics can provide a rough
proxy. Language statistics is useful in that it is part of the inputs to
everyday experiences (through spoken and written language), may afford
many of the `experiences' that people have with infrequently encountered
items, like cows or helicopters, and provides accessible measures.
Previous work also suggests that using language statistics (such as
lexical co-occurrence or cosine distance of word embeddings) as a proxy
can be a good predictor for similarity reasoning. Griffiths, Steyvers,
\& Tenenbaum (2007) showed that a model trained on word-document
co-occurrence can predict word association and the effects of semantic
association on a variety of linguistic tasks. Lexical semantic models
that are constructed using lexical co-occurrence (in comparison to
annotated relations) have been shown to perform well on predicting human
judgments about similarity between word pairs that are thematically or
taxonomically related (Rohde, Gonnerman, \& Plaut, 2006). Word
embeddings like fastText have been demonstrated to be good predictors
for similarity judgments (Liu, Feng, Wu, Chan, \& Fulton (2019),
Jatnika, Bijaksana, \& Suryani (2019)).

Our study uses cosine distances of fastText word vectors as a measure of
lexical co-occurrence\footnote{We also carried out our analysis using
  raw lexical co-occurrences and obtained similar results.}. fastText is
a system that uses lexical co-occurrence information to generate a
vector representing each word in its lexicon (Mikolov, Grave,
Bojanowski, Puhrsch, \& Joulin, 2018). fastText has also been shown to
be sensitive towards cultural effects on word meanings: Thompson \&
Lupyan (2020) showed that the distribution of semantic meaning clusters
generated by fastText trained on language-specific corpora correlates to
the cultural, historical and geographical distances of such languages
languages. This is a proof of concept for our use of fastText in
different languages with varying levels of relatedness.

We note that language statistics can incorporate both cultural-specific
environmental statistics (the inputs of similarity judgment) and
taxonomic/thematic preferences (the conceptualization of similarity).
For example, cultural features (farming) can lead to differences in
environmental statistics (seeing cows and grass) and in turn these can
influence language (talking more about cows and grass). But cultural
features (preferring thematic relations) could also more directly cause
individuals to talk differently about the same experiences (mentioning
what cows eat rather than what other animals cows are like). Therefore,
our approach looks at the extent to which language statistics can
predict cross-cultural differences in similarity judgment with the
understanding that language statistics is a proxy for both inputs and
conceptualization of similarity judgment\footnote{However, in Q3, we
  investigate whether differing culture-based conceptualization of
  similarity has unique contribution to cross-cultural variation of
  similarity jdugment.}.

@@

Our study tests the hypothesis that, rather than (or in addition to) the
variation in conceptualization of similarity across cultural groups,
variation in the input to similarity judgment (namely, environmental
statistics as proxied by language statistics) can be used to predict
cross-cultural differences observed in how people evaluate similarity.
We note that our study is correlational and cannot make a causal claim
about the relationship of environmental statistics to similarity
judgment, but it may be possible to gain traction on potential
mechanisms by examining whether variation in similarity judgments covary
with environmental statistics that differ across cultural and linguistic
contexts. Our specific research questions are as follows:~

Q1: Do we replicate Ji et al. (2004) and extend the results to another
East Asian culture, namely Vietnam? Vietnam is a Southeast Asian country
that borders China and is historically greatly influenced by Chinese
culture (Hui, 2002). Therefore, it serves as a suitable cultural context
to investigate whether the claim made by Ji et al. (2004) and previous
studies, that Eastern and Western cultures have different notions of
similarity, extends beyond mainland China.

Q2: To what extent do cross-context environmental statistics (as proxied
by language) align with variation in similarity judgment?

Q3: Is there evidence for a unique contribution from conceptualization
of similarity?

Q4: Is language statistics viable as a standalone predictor, or is it
simply measuring conceptualization of similarity in a different way?~

To preview our results, we find that, while we do not find the predicted
Eastern/Western contrast in conceptualization of similarity in our
Vietnam extension, we do find that both language-specific statistics and
cultural contexts are a good predictor for cross-cultural differences in
similarity judgments. Language statistics are also a good predictor for
filler items with no taxonomic/thematic structure. Our findings are in
contrast to previous accounts, by which culture induces differing
conceptions of similarity (varying between taxonomic and thematic).
these findings provide an alternative, and more specific, account by
which language may explain these cross-cultural differences without
invoking variation in notions of similarity.

@@

We excluded 8 US participants, 62 Vietnam participants, and 16 China
participants who missed 2 or more attention checks. We followed 4
exclusion criteria that aim to retain only participants who are
influenced by one culture: (1) non-native speakers of English and
Vietnamese, respectively, (2) fluent in at least one of the other two
study languages (Vietnamese for US participants, English for Vietnamese
participants and Chinese participants), (3) have lived outside of the
test country (US, Vietnam, or China) for more than two years, and (4)
have significant international experience (more than 6 international
experiences of 2 days or longer.) We did not use a particular criterion
for a language if it would exclude 25\% or more of any one sample. In
this round of exclusion, we excluded 73 US, 27 Vietnam participants, and
35 China participants. After these exclusions, the US sample included
119 participants (30M, 84F, 3 non-binary, 2 other), with mean age = 22.2
(SD = 8.15) and median age = 20. The Vietnam sample included 110
participants (34M, 71F, 5 other), with mean age = 22.21 (SD = 5.81) and
median age = 21. The China sample included 149 participants (61M, 87F, 1
other), with mean age = 23.1 (SD = 3.65) and median age = 23.\footnote{A
  table summarizing number of participants lost at each round of
  exclusions is included in the Supplementary Information.}

We preregistered a more stringent exclusion criterion where participants
were excluded if they missed any attention checks. However, this led to
a small sample size, especially for Vietnam context (US = 109, China =
132, Vietnam = 57). Our reported results with the less stringent
exclusion criterion (detailed above) is largely not different from the
results with the preregistered criterion. Any differences are noted in
the Analysis section.

\hypertarget{stimuli-materials}{%
\subsection{Stimuli Materials}\label{stimuli-materials}}

We adapted stimuli from previous studies to create a set of test triads
consisting of a cue, with one thematic and one taxonomic match option.
For example, ``cow,'' ``grass,'' and ``chicken,'' where ``cow'' is the
cue, ``grass'' is the thematic match, and ``chicken'' the taxonomic
match. We included 105 such triads, a superset including triads pulled
from supplemental information and in-text examples across the
literature, and others that we adapted or created . We selected triads
on the basis of cultural familiarity in the US, Vietnam, and China. The
triads were originally in English; they were translated to Vietnamese
and Mandarin by a fluent bilingual speaker in each language. The
translations were then checked for accuracy after backtranslation to
English by another fluent bilingual in each language who was naive to
the original English versions. All materials are included in the
Supplementary Information.

\hypertarget{procedure}{%
\subsection{Procedure}\label{procedure}}

Each participant completed all 105 triads in sets of 21 trials at a time
(10 test triads, 10 filler triads, and 1 attention check per page), by
selecting the match most related to the cue (``Which thing is most
closely related to the bolded item?'' and translated equivalents). The
test triads were presented with 105 filler triads mixed in, to obscure
the taxonomic-thematic two-answer forced choice structure of the test
stimuli and reduce the likelihood that participants would become aware
of the design. The filler triads were groups of three semantically
related words, but where the match options were not distinguished by
thematic vs.~taxonomic similarity, e.g., cue ``bird'' with match options
``lizard'' and ``toad.'' Additionally, we included 10 attention check
trials, which were formatted like the test and filler triads but
included an instruction instead of a cue item, e.g., ``Choose wife''
with match options ``wife'' and ``husband.'' In total, each participant
completed 210 similarity judgments and 10 attention check questions,
with triads presented in randomized orders that varied between subjects.

@@

Our general approach is to predict behavioral preferences in similarity
judgment using relative similarity between word embeddings.

@@

We use the fastText pre-trained models of English, Mandarin, and
Vietnamese in Grave, Bojanowski, Gupta, Joulin, \& Mikolov (2018). These
models are trained on Common Crawl and Wikipedia using We distribute
pre-trained word vectors for 157 languages, trained on Common Crawl and
Wikipedia using fastText. These models were trained using a Continuous
Bag of Words (CBOW) with position-weights and a window of size 5. The
models use character n-grams of length 5 and 10 negative examples. From
the aforementioned models, we retrieve the word vectors (dimension 300)
for each word we are interested in.

@@

To give an intuition for our model, consider again the cow-grass-chicken
triad: we retrieved word vectors for ``cow'' and ``grass'', and
calculate the cosine distance between these vectors. Similarly, we
retrieved vectors for ``cow'' and ``chicken'' and calculate the cosine
distance between them. Our similarity prediction is then inversely
proportional to the ratio of cosine distance of these pairs. This is
because a larger cosine distance means the word vectors are further
apart, and thus the words are less similar. For example, if the cosine
distance of thematic cow-grass is 0.7 and the cosine distance of
taxonomic cow-chicken is 0.3, then our model predicts, correspondingly,
that 30\% of responses to the triad will be grass, and the other 70\%
chicken.

In practice, we calculated the cosine distance between each cue-thematic
match (thematic cosine distance) and cue-taxonomic match (taxonomic
cosine distance), using the spatial.distance.cosine function from the
SciPy package (Virtanen et al., 2020). We then calculated the thematic
cosine distance proportion as thematic cosine distance over the sum of
taxonomic cosine distance and thematic cosine distance. We did this for
all three corpora. We were able to obtain predictions for all triads in
all languages. We then use a mixed-effects regression to evaluate how
well each corpus model predicts participants' similarity judgments,
across triads and cultural contexts.

@@

\hypertarget{replication-of-previous-work-and-extension-to-a-vietnamese-sample}{%
\subsection{1. Replication of previous work and extension to a
Vietnamese
sample}\label{replication-of-previous-work-and-extension-to-a-vietnamese-sample}}

Following previous work, we would expect participants from mainland
China to prefer thematic matches more than US participants (to use the
cow-grass-chicken triad: we would expect participants from mainland
China to prefer the cow-grass match over the cow-chicken match to a
larger extent compared to US participants). We would also expect that
participants from Vietnam would pattern with China, and also show a
significantly higher preference for thematic matches compare to US
participants.

@@

To test for cross-context differences in similarity judgments between
the countries, we ran a mixed-effects logistic regression predicting
triad responding (taxonomic or thematic) with country (US, China, or
Vietnam) as a fixed effect. As random effects, we included an intercept
per subject and one per triad, as well as by-triad random slopes for
country to account for variation in the country effect across triads.

@@

\hypertarget{language-statistics-as-a-predictor-for-cross-cultural-variation-in-similarity-judgments}{%
\subsection{2. Language statistics as a predictor for cross-cultural
variation in similarity
judgments}\label{language-statistics-as-a-predictor-for-cross-cultural-variation-in-similarity-judgments}}

\hypertarget{single-corpus-model}{%
\subsubsection{Single corpus model}\label{single-corpus-model}}

To test whether variation in language statistics can explain differences
in similarity judgments between US and Vietnam participants, we compare
logistic mixed-effects regression models fit to the thematic responding
data from each country separately. We first ask how well each corpus
model (English, Vietnamese, or Mandarin) predicts similarity judgments
by speakers of the corresponding language (US, Vietnam, or China). To do
this, we use a mixed-effects logistic regression to predict triad
responses (0=taxonomic or 1=thematic) with corpus prediction (proportion
of cosine distance) as a fixed effect and participant and triad as
random effects. If environmental statistics (as proxied by language
statistics) contribute to the differences in similarity judgments, we
would expect each language corpus to be a good predictor for similarity
judgment responding in its corresponding context.

We found that all corpora are significant predictors of all cultural
context responding, with p \textless{} 0.05 and \(\beta\) from -8.69 to
-2.4. (For a full report, see Supplementary Information.)

While each corpus is a good predictor for its corresponding context, the
fact that all corpora covary with all cultural contexts suggests that
language is such a rich proxy of human experience that even using the
wrong proxy (e.g.~Mandarin for US responding) is informative. A single
corpus model might therefore not be informative in culture-specific
ways, but might only reflect consistency in experiences across cultures.

\hypertarget{multiple-corpora-model}{%
\subsubsection{Multiple corpora model}\label{multiple-corpora-model}}

If language statistics is able to predict meaningful culture-specific
variation in similarity judgment (rather than just consistency across
cultures), we would expect each corpus to be the best predictor of its
corresponding culture compared to the other two corpora. We directly
compare the corpus models by including both as fixed effects in three
mixed-effect regressions (predicting US, Vietnam and China responding)
with the same random effects as above.

For US responding: only the English (EN) corpus is a significant
predictor\footnote{With preregistered exclusion criterion: only the
  English (EN) and Mandarin (ZH) corpus are significant predictors.}. EN
corpus: \(\beta\) = -6.96, \(\chi^2\)(1) = 16.57, p \textless{} .001. VI
corpus: \(\beta\) = -2.22, \(\chi^2\)(1) = 3.73, p = 0.054. ZH corpus:
\(\beta\) = -3.18, \(\chi^2\)(1) = 3.37, p = 0.066.

For Vietnam responding: only the Vietnamese (VI) and Mandarin (ZH)
corpus are significant predictors\footnote{With preregistered exclusion
  criterion: all three corpora are significant predictors.}. EN corpus:
\(\beta\) = -2.98, \(\chi^2\)(1) = 2.58, p = 0.108. VI corpus: \(\beta\)
= -2.75, \(\chi^2\)(1) = 4.84, p = 0.028. ZH corpus: \(\beta\) = -4.39,
\(\chi^2\)(1) = 5.49, p = 0.019.

For China responding: only the Mandarin (ZH) and English (EN) corpus are
significant predictors. EN corpus: \(\beta\) = -3.32, \(\chi^2\)(1) =
7.18, p = 0.007. VI corpus: \(\beta\) = -1.59, \(\chi^2\)(1) = 3.63, p =
0.057. ZH corpus: \(\beta\) = -5.31, \(\chi^2\)(1) = 17.69, p
\textless{} .001.

We observed some level of language specificity from this analysis. The
English corpus is the best predictor for US responding, and the Mandarin
corpus is the best predictor for China response. While this is not the
case with the Vietnamese corpus and the Vietnam responding, the
Vietnamese corpus is still a significant predictor for the Vietnam
responding (Figure 2). These results strengthens our hypothesis that
variation in inputs to similarity judgment (as proxied by language
statistics) can predict cross-cultural variations of similarity
judgment.

@@

However, in all cultural contexts, adding the other two corpora produces
a significantly better fit than the identical model without the
additional corpora, and only the corresponding corpus included as a
predictor (US response: \(\chi^2\)(2) = 7.93, p = 0.019; Vietnam
response: \(\chi^2\)(2) = 13.72, p = 0.001; China response:
\(\chi^2\)(2) = 10.56, p = 0.005). This analysis suggests that
culture-specific inputs to similarity judgment (as proxied by language
statistics) do not fully explain cross-cultural differences in
similarity judgment.
